<!-- Material and methods is the text of the Methods. Each analysis lives in
     the results section that reports it, beside the numbers it gives, and
     the file is named after it: 04.1_richness.qmd, 04.2_composition.qmd.
     What every analysis needs is prepared here, once, in the chunks below:
     the packages, the data read from data/ -- never from where the
     originals are kept -- and the helpers the text uses. Every cleaning
     step is code a reviewer can read. They print nothing: _quarto.yml sets
     include: false for every chunk, and only figures and tables turn it
     back on.
     Packages are loaded in a chunk of their own, without cache: a cached
     chunk would not load them again. -->

```{r}
#| label: packages
library(vegan)
```

```{r}
#| label: data
#| cache: true
# The two tables, read from data/ with paths relative to the project root.
abund <- read.csv(here::here("data/tree_abundance.csv"), check.names = FALSE)
env   <- read.csv(here::here("data/plot_environment.csv"), check.names = FALSE)

# The plots in the same order in both tables, or nothing below is right.
stopifnot(identical(abund$plot, env$plot))

Y_counts <- as.matrix(abund[, -1])
species  <- colnames(Y_counts)
habitat  <- factor(env$Habitat)

n_plots   <- nrow(Y_counts)
n_species <- ncol(Y_counts)
n_ind     <- sum(Y_counts)
richness  <- rowSums(Y_counts > 0)
n_hab     <- table(habitat)
n_top     <- min(30L, n_species)   # the species listed in the supplement
# The permutations of every test: what a paper reports. The Methods say how
# many, and the analyses of composition run them.
n_perm <- 999

# How the habitats are named in the text ("old forest on slopes") and in the
# tables and axes ("Old, slope"). A level not in the list keeps its name.
hab_text <- function(x) {
  lab <- c(Young = "young forest", OldLow = "old forest on low ground",
           OldSlope = "old forest on slopes", OldHigh = "old forest on high ground",
           Swamp = "swamp")
  ifelse(x %in% names(lab), lab[x], x)
}
hab_title <- function(x) {
  lab <- c(Young = "Young", OldLow = "Old, low", OldSlope = "Old, slope",
           OldHigh = "Old, high", Swamp = "Swamp")
  ifelse(x %in% names(lab), lab[x], x)
}
# A P-value as the text writes it: "= 0.012", or "< 0.001".
fmt_p <- function(p) {
  if (p < 0.001) "< 0.001" else paste("=", formatC(p, digits = 3, format = "f"))
}
# The part of a species name that goes in italics -- the genus and the
# epithet, not "sp. A" -- and the part that stays in roman.
sp_italic <- function(x) sub("( sp\\..*| complex)$", "", x)
sp_roman  <- function(x) substring(x, nchar(sp_italic(x)) + 1L)
sp_label  <- function(x) bquote(italic(.(sp_italic(x))) * .(sp_roman(x)))
```

## Data

We re-analysed the trees of the 50-ha forest dynamics plot of Barro Colorado
Island, Panama [@Condit2002], distributed as the `BCI` dataset of the R
package vegan [@R-vegan]: the number of trees of at least 10 cm in diameter at
breast height of each of `r n_species` species in each of `r n_plots` plots of
1 ha, censused in 1982, and the dominant habitat of each plot. The plots are
unevenly spread among the `r nlevels(habitat)` habitats
(`r knitr::combine_words(paste(n_hab, hab_text(names(n_hab))))`). The two
tables are in `data/` and are described, column by column, in
`data/metadata/`.

## Statistical analysis

Species richness, the number of species recorded in a plot, was modelled with
a generalized linear model (GLM) with a Poisson distribution and habitat as
the predictor, tested with a likelihood-ratio test.

The composition of the assemblages was compared with Bray-Curtis
dissimilarities between plots, calculated on square-root-transformed
abundances. Its difference among habitats was tested with a permutational
multivariate analysis of variance [PERMANOVA, @Anderson2001], and the
homogeneity of the dispersions of the habitats with a permutational test of
multivariate dispersions [PERMDISP, @Anderson2006], each with `r n_perm`
permutations. The dissimilarities were represented with a non-metric
multidimensional scaling (NMDS) in two dimensions. The analyses of composition
were run with vegan [@R-vegan], and all analyses in R [@R-base].
