<!-- The analysis lives in the section that reports it. The chunks below read
     the data from data/ -- never from where the originals are kept -- and
     every cleaning step is code a reviewer can read. They print nothing:
     _quarto.yml sets include: false for every chunk, and only figures and
     tables turn it back on.
     #| cache: true keeps a slow chunk from running at every render. When the
     data underneath it change, run clean_cache(): a change in data/ is the
     one thing knitr's cache does not notice.
     Packages are loaded in a chunk of their own, without cache: a cached
     chunk would not load them again. -->

```{r}
#| label: packages
library(vegan)
```

```{r}
#| label: data
#| cache: true
# The two tables, read from data/ with paths relative to the project root.
abund <- read.csv(here::here("data/tree_abundance.csv"), check.names = FALSE)
env   <- read.csv(here::here("data/plot_environment.csv"), check.names = FALSE)

# The plots in the same order in both tables, or nothing below is right.
stopifnot(identical(abund$plot, env$plot))

Y_counts <- as.matrix(abund[, -1])
species  <- colnames(Y_counts)
habitat  <- factor(env$Habitat)

n_plots   <- nrow(Y_counts)
n_species <- ncol(Y_counts)
n_ind     <- sum(Y_counts)
richness  <- rowSums(Y_counts > 0)
n_hab     <- table(habitat)
n_top     <- min(30L, n_species)   # the species listed in the supplement

# How the habitats are named in the text ("old forest on slopes") and in the
# tables and axes ("Old, slope"). A level not in the list keeps its name.
hab_text <- function(x) {
  lab <- c(Young = "young forest", OldLow = "old forest on low ground",
           OldSlope = "old forest on slopes", OldHigh = "old forest on high ground",
           Swamp = "swamp")
  ifelse(x %in% names(lab), lab[x], x)
}
hab_title <- function(x) {
  lab <- c(Young = "Young", OldLow = "Old, low", OldSlope = "Old, slope",
           OldHigh = "Old, high", Swamp = "Swamp")
  ifelse(x %in% names(lab), lab[x], x)
}
# A P-value as the text writes it: "= 0.012", or "< 0.001".
fmt_p <- function(p) {
  if (p < 0.001) "< 0.001" else paste("=", formatC(p, digits = 3, format = "f"))
}
# The part of a species name that goes in italics -- the genus and the
# epithet, not "sp. A" -- and the part that stays in roman.
sp_italic <- function(x) sub("( sp\\..*| complex)$", "", x)
sp_roman  <- function(x) substring(x, nchar(sp_italic(x)) + 1L)
sp_label  <- function(x) bquote(italic(.(sp_italic(x))) * .(sp_roman(x)))
```

## Data

We re-analysed the trees of the 50-ha forest dynamics plot of Barro Colorado
Island, Panama [@Condit2002], distributed as the `BCI` dataset of the R
package vegan [@R-vegan]: the number of trees of at least 10 cm in diameter at
breast height of each of `r n_species` species in each of `r n_plots` plots of
1 ha, censused in 1982, and the dominant habitat of each plot. The plots are
unevenly spread among the `r nlevels(habitat)` habitats
(`r knitr::combine_words(paste(n_hab, hab_text(names(n_hab))))`). The two
tables are in `data/` and are described, column by column, in
`data/metadata/`.

## Statistical analysis

```{r}
#| label: richness-model
#| cache: true
m_rich  <- glm(richness ~ habitat, family = poisson)
lrt     <- anova(m_rich, test = "Chisq")
chi_hab <- lrt["habitat", "Deviance"]
df_hab  <- lrt["habitat", "Df"]
p_hab   <- lrt["habitat", "Pr(>Chi)"]
dispersion <- sum(residuals(m_rich, type = "pearson")^2) / df.residual(m_rich)

# The expected richness of each habitat, with its 95% confidence interval.
nd   <- data.frame(habitat = factor(levels(habitat), levels = levels(habitat)))
pr   <- predict(m_rich, newdata = nd, type = "link", se.fit = TRUE)
pred_rich <- data.frame(habitat = levels(habitat),
                        fit   = exp(pr$fit),
                        lower = exp(pr$fit - 1.96 * pr$se.fit),
                        upper = exp(pr$fit + 1.96 * pr$se.fit))
hab_max <- pred_rich$habitat[which.max(pred_rich$fit)]
hab_min <- pred_rich$habitat[which.min(pred_rich$fit)]
```

```{r}
#| label: composition-model
#| cache: true
# 999 permutations: what a paper reports. They take a few seconds, once:
# the chunk is cached.
n_perm <- 999
# Bray-Curtis dissimilarities between plots, on square-root abundances, so
# that the few very abundant species do not decide everything.
Y_sqrt <- sqrt(Y_counts)
d_bray <- vegdist(Y_sqrt, method = "bray")

# PERMANOVA: does the composition differ among habitats?
perm     <- adonis2(d_bray ~ habitat, permutations = n_perm, by = "terms")
tab_perm <- as.data.frame(perm)
r2_hab   <- tab_perm["habitat", "R2"]
f_hab    <- tab_perm["habitat", "F"]
pp_hab   <- tab_perm["habitat", "Pr(>F)"]

# PERMDISP: do the habitats differ in how variable their composition is?
disp   <- betadisper(d_bray, habitat)
pdisp  <- permutest(disp, permutations = n_perm)
f_disp <- pdisp$tab["Groups", "F"]
p_disp <- pdisp$tab["Groups", "Pr(>F)"]

# NMDS of the same dissimilarities.
nmds <- metaMDS(Y_sqrt, distance = "bray", k = 2, trymax = 50,
                autotransform = FALSE, trace = FALSE)

# The six most abundant species, named on the ordination.
top_sp <- species[order(colSums(Y_counts), decreasing = TRUE)[1:6]]
```

Species richness, the number of species recorded in a plot, was modelled with
a generalized linear model (GLM) with a Poisson distribution and habitat as
the predictor, tested with a likelihood-ratio test (dispersion
`r round(dispersion, 2)`).

The composition of the assemblages was compared with Bray-Curtis
dissimilarities between plots, calculated on square-root-transformed
abundances. Its difference among habitats was tested with a permutational
multivariate analysis of variance [PERMANOVA, @Anderson2001], and the
homogeneity of the dispersions of the habitats with a permutational test of
multivariate dispersions [PERMDISP, @Anderson2006], each with `r n_perm`
permutations. The dissimilarities were represented with a non-metric
multidimensional scaling (NMDS) in two dimensions. The analyses of composition
were run with vegan [@R-vegan], and all analyses in R [@R-base].
